---
title: "GitHub-style waffle plots in R"
author: "Matti Vuorre"
date: 2016-03-24
categories : ["data science"]
tags: ["R", "data visualization", "tutorial"]
description: "Plotting activity over time in R with ggplot2."
draft: false
---

```{r, echo=F, warning=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(
    fig.retina = 2,
    fig.align = 'center', 
    comment = "#", 
    warning = F, 
    message = F, 
    cache = T,
    fig.width=9,
    fig.height=6
)
```

In this post, I'll show how to create GitHub style "waffle" plot in [R](https://cran.r-project.org/) with the [ggplot2](http://had.co.nz/ggplot2/book) plotting package. 

# Simulate activity data

First, I'll create a data frame for the simulated data, initializing the data types: 

```{r gen_dataframe}
library(dplyr)
d <- data_frame(
    date = as.Date(1:813, origin = "2014-01-01"),
    year = format(date, "%Y"),
    week = as.integer(format(date, "%W")) + 1,  # Week starts at 1
    day = factor(weekdays(date, T), 
                 levels = rev(c("Mon", "Tue", "Wed", "Thu",
                                "Fri", "Sat", "Sun"))),
    hours = 0)
```

And then simulate hours worked for each date. I'll simulate hours worked separately for weekends and weekdays to make the resulting data a little more realistic, and also simulate missing values to data (that is, days when no work occurred).

```{r}
set.seed(1)
# Simulate weekends
weekends <- filter(d, grepl("S(at|un)", day))
# Hours worked are (might be) poisson distributed
weekends$hours <- rpois(nrow(weekends), lambda = 4)
# Simulate missing days with probability .7
weekends$na <- rbinom(nrow(weekends), 1, 0.7)
weekends$hours <- ifelse(weekends$na, NA, weekends$hours)

# Simulate weekdays
weekdays <- filter(d, !grepl("S(at|un)", day))
weekdays$hours <- rpois(nrow(weekdays), lambda = 8)  # Greater lambda
weekdays$na <- rbinom(nrow(weekdays), 1, 0.1)  # Smaller p(missing)
weekdays$hours <- ifelse(weekdays$na, NA, weekdays$hours)

# Concatenate weekends and weekdays and arrange by date
d <- bind_rows(weekends, weekdays) %>% 
    arrange(date) %>%  # Arrange by date
    select(-na)  # Remove na column
d
```

# Waffle-plot function

Then I'll create a function that draws the waffle plot. If you have similarly structured data, you can copy-paste the function and use it on your data (but make sure the following packages are installed.)

```{r viz_func}
library(ggplot2)
library(viridis)  # Color palette
library(ggthemes)
gh_waffle <- function(data, pal = "D", dir = -1){
    
    p <- ggplot(data, aes(x = week, y = day, fill = hours)) +
        scale_fill_viridis(name="Hours", 
                           option = pal,  # Variable color palette
                           direction = dir,  # Variable color direction
                           na.value = "grey93",
                           limits = c(0, max(data$hours))) +
        geom_tile(color = "white", size = 0.4) +
        facet_wrap("year", ncol = 1) +
        scale_x_continuous(
            expand = c(0, 0),
            breaks = seq(1, 52, length = 12),
            labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                       "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
        theme_tufte(base_family = "Helvetica") +
        theme(axis.title = element_blank(),
              axis.ticks = element_blank(),
              legend.position = "bottom",
              legend.key.width = unit(1, "cm"),
              strip.text = element_text(hjust = 0.01, face = "bold", size = 12))
    
    print(p)
}
```

## Using the waffle plot function

`gh_waffle()` takes three arguments, the first, `data` is a data frame with columns `date` (type: Date), `year` (number or character), `week` (number), `day` (an ordered factor to make days run from top to bottom on the graph), and `hours` (number). The second option to `gh_waffle()`, `pal` specifies one of four color palettes used by the `viridis` color scale, and can be `"A"`, `"B"`, `"C"`, or `"D"`. The default is "D", which is also what GitHub uses. The last option, `dir` specifies the direction of the color scale, and can be either `-1` or `1`. The GitHub default is -1. 

Using `gh_waffle()` with the default settings, only providing the data frame `d`, gives the following result:

```{r viz}
gh_waffle(d)
```

Here's the same plot with the three other color palettes:

```{r viz2}
for (pal in c("A", "B", "C")) {
    gh_waffle(d, pal)
}
```

Unless you're plotting some really awful events, I think it's best to stick with the default color palette. That's it for today folks. 

# Further reading & references

* [Faceted heatmaps with ggplot2](http://rud.is/b/2016/02/14/making-faceted-heatmaps-with-ggplot2/) (Inspiration for this post.)
* [dplyr](https://CRAN.R-project.org/package=dplyr)
* [ggplot2](http://had.co.nz/ggplot2/book)
* [viridis](https://github.com/sjmgarnier/viridis)
* [ggthemes](https://github.com/jrnold/ggthemes)
