---
title: SIPS 2017 Conference Tweets
author: Matti Vuorre
date: '2017-08-03'
slug: sips-2017-conference-tweets
categories:
  - psychology
tags:
  - data visualization
  - psychology
  - SIPS
output:
  blogdown::html_page:
    toc: true
    number_sections: false
    toc_depth: 1
draft: no
bibliography: "`r here::here('static/bibliography/blog.bib')`"
summary: "Exploring SIPS Tweets with R."
---

```{r setup, include=F}
library(rtweet)
library(tidytext)
library(tidyverse)
theme_set(theme_linedraw())
knitr::opts_chunk$set(error = T, echo = F, message = F, cache = T)
```

```{r save-data, eval = F}
tweets <- search_tweets("#SIPS2017", 5000)
save(tweets, file = "static/data/sips-2017-twitter.rda")
```

The second annual meeting of the Society for the Improvement of Psychological Science ([SIPS](https://osf.io/jtcu9/)) is over, sadly. The event was actively tweeted with [#SIPS2017](https://twitter.com/hashtag/SIPS2017), so we can use R to explore the tweets related to the conference. In this blog post, I do a small retrospective on the event by accessing and visualizing the SIPS tweets through Twitter's API.

# Obtaining Data

"Accessing an API" sounds like a big deal, but fortunately [there's an R package for that](https://www.youtube.com/watch?v=yhTerzNFLbo). I used the **rtweet** package [@rtweet-package] to download 2829 tweets with the #SIPS2017 hashtag. I saved the resulting data, which can be loaded with the following R command:

```{r load-data, echo = T}
load(here::here("static/data/sips-2017-twitter.rda"))
```

# Timeline of Tweets

The timeline of tweets reveals that SIPSers were most active on Twitter on Monday, and there's a clear spike of activity at 6pm on Monday.

```{r, fig.height = 3}
library(lubridate)
days <- c("Friday", "Saturday", "Sunday", "Monday", "Tuesday", "Wednesday")
tweets %>% 
    mutate(datetime = as_datetime(created_at, tz = Sys.timezone()),
           hour = hour(datetime),
           weekday = wday(datetime, label = T, abbr = F)) %>%
    filter(weekday %in% days) %>%
    mutate(weekday = factor(weekday, levels = days)) %>% 
    group_by(date = as_date(datetime), weekday, hour) %>%
    count(sort = T) %>%
    ggplot(aes(hour, n)) +
    geom_bar(stat = "identity", col = "black", fill = "black") +
    labs(x="Time", y="Tweets", title="#SIPS2017 Twitter activity") +
    facet_wrap("weekday", nrow = 1) +
    coord_cartesian(expand = 0, ylim = c(0, 250)) +
    theme(panel.spacing = unit(0, "cm"))
```

# Top tweets

Here are the top 5 SIPS non-retweets.

```{r, eval = T, results = 'asis'}
tweets %>% 
    filter(!is_retweet) %>% 
    arrange(desc(retweet_count)) %>% 
    rename(retweets = retweet_count) %>% 
    select(screen_name, created_at, retweets, text) %>% 
    slice(1:5) %>% 
    mutate(text = gsub("@", "", text)) %>% 
    knitr::kable(format = "html")
```

# Top Hashtags

Here are the top hashtags people used with the SIPS hashtag.

```{r}
tweets %>% 
    unnest_tokens(tags, hashtags) %>% 
    count(tags, sort = T) %>% 
    filter(n > 1, tags != "sips2017") %>% 
    ggplot(aes(n, reorder(tags, n) )) +
    geom_point() +
    labs(title = "Most tweeted tags at SIPS 2017",
         x="Number of tweets", y="Hashtag") 
```

# Who mentioned who in their tweets?

And finally the obligatory graph: Arrows represent tweets going from someone to someone else.

```{r, eval = T, warning = F, fig.width = 12, fig.height = 12}
library(igraph)
g <- tweets %>% 
    filter(!is.na(mentions_screen_name)) %>% 
    select(screen_name, mentions_screen_name) %>% 
    mutate(mentions_screen_name = strsplit(mentions_screen_name, " ")) %>% 
    unnest(mentions_screen_name) %>% 
    graph_from_data_frame(directed = T)
V(g)$label <- V(g)$name
V(g)$id <- V(g)$name

plot(g,
     margin = -.12,
     layout = layout_on_sphere(g),
     edge.color = adjustcolor("dodgerblue1", .25),
     edge.arrow.size = .5,
     edge.width = 1.1,
     vertex.color = adjustcolor("dodgerblue1", .1),
     vertex.frame.color=adjustcolor("dodgerblue1", .2),
     vertex.size = 2,
     vertex.label.family = "Helvetica",
     vertex.label.font = 1,
     vertex.label.color = "black",
     vertex.label.cex = sqrt(1.5+degree(g))/8,
     vertex.label = ifelse(degree(g, mode = "in") > 30, V(g)$label, NA))
```



You'll find the full source code of this blog post on my [GitHub page](https://github.com/mvuorre/blog).

# References
